statistical-analysis

### 2025-04-08 assets/statistical-analysis.qmd
Edit manuscript

```{r}
#load needed packages. make sure they are installed.
library(here) #for data loading/saving
library(dplyr)
library(skimr)
library(ggplot2)
install.packages("maps")
library(maps)
install.packages("ggcorrplot")
library(ggcorrplot)
install.packages("plm")
library(plm)
install.packages("gt")
library(gt)
install.packages("car")
library(car)
install.packages("MASS")  # Jika belum terinstall
library(MASS)
install.packages("tidyverse")
library(tidyverse)
install.packages("reshape2")  # Install the package (only needed once)
library(reshape2)  # Load the package
install.packages("scales")
library(scales)

library(splines)
library(here)
library(knitr)
install.packages("kableExtra")
library(kableExtra)
install.packages("GGally")
library(GGally)
# Install and load the patchwork package
install.packages("patchwork")
library(patchwork)

install.packages("gridExtra")  # If not installed
library(gridExtra)

install.packages("lmerTest")
library(lmerTest)
install.packages("stargazer")
library(stargazer)

@@

asia_africa_location <- here("data", "processed-data", "asia_africa.rds")
asia_africa <- readRDS(asia_africa_location)

# In the study, I will focus on Water and Sanitation and remove Lesotho as it considered as outlier
asia_africa <- asia_africa%>%
  filter(country != "Lesotho") %>% #Drop Lesotho from the observation
  select("tbi", "year", "pro_basicwater", "pro_safewater", "pro_basicsanitation", "pro_safesanitation", "hdi", "country", "region")

colnames(asia_africa)
```

@@

***Test the linearity***

```{r}
# Adding Fixed-Effect to Make Conditional Models ---
ggplot(asia_africa, aes(x = year, y = tbi)) + 
  geom_line(aes(group=country), col="grey") + 
  stat_smooth(method="loess", col="black", lwd=1.5, se=FALSE)+
  scale_x_continuous(name = "years") + 
  scale_y_continuous(name = "Tuberculosis Incidence") +
  theme_bw() + 
  theme(axis.text=element_text(size=14, colour="black"), 
        axis.title=element_text(size=14,face="bold")) +
  theme(strip.text.x = element_text(size = 14))+ theme(legend.position="none") 
```

```{r}
# Convert dataset to long format for faceting
asia_africa_long <- asia_africa %>%
  pivot_longer(cols = c(pro_basicwater, pro_safewater,
                        pro_basicsanitation, pro_safesanitation, hdi),
               names_to = "variable", values_to = "value")

# Create scatter plot faceted by independent variable
scatter_faceted <- ggplot(asia_africa_long, aes(x = value, y = tbi, color = country)) + 
  geom_point(alpha = 0.6) +  # Scatter points, colored by country
  stat_smooth(method = "loess", col = "black", lwd = 1.2, se = FALSE) +  # LOESS (non-linear) in black
  facet_wrap(~ variable, scales = "free_x") +  # Facet by variable
  labs(x = "Independent Variable Value", y = "Tuberculosis Incidence (TBI)") +
  theme_bw() + 
  theme(axis.text = element_text(size = 12, colour = "black"), 
        axis.title = element_text(size = 14, face = "bold"),
        strip.text = element_text(size = 14, face = "bold"), 
        legend.position = "none")  # Hides the legend

nonrelationshiplot_location <- here( "code","analysis-code", "figures", "nonlinear.png") # to set up location for the pictures created 
ggsave(filename = nonrelationshiplot_location, plot=scatter_faceted, width = 12, height = 12, units = "in", dpi = 300) # save the pictures created 
# Show plot
print(scatter_faceted)

@@

Non-linearity test 
```{r}
# Install package if not already installed
install.packages("lmtest")
library(lmtest)

# Fit a linear model
lm_model <- lm(tbi ~ pro_cleanfuel + pro_basicwater + pro_safewater + 
               mean_exposure_pm2.5 + total_greenhouses + 
               pro_basicsanitation + pro_safesanitation + gdp + hdi, 
               data = asia_africa)

# Run the RESET test for linearity
resettest(lm_model, power = 2:3)

@@

# Fit a linear model (LM)
lm_model <- lm(tbi ~ pro_cleanfuel + pro_basicwater + pro_safewater + 
               mean_exposure_pm2.5 + total_greenhouses + 
               pro_basicsanitation + pro_safesanitation + gdp + hdi, 
               data = asia_africa)

# Fit a non-linear model (GAM)
gam_model <- gam(tbi ~ s(pro_cleanfuel) + s(pro_basicwater) + s(pro_safewater) + 
                  s(mean_exposure_pm2.5) + s(total_greenhouses) + 
                  s(pro_basicsanitation) + s(pro_safesanitation) + 
                  s(gdp) + s(hdi), 
                 data = asia_africa, method = "REML")

# Compare AIC values
AIC(lm_model, gam_model)

# Plot residuals vs. fitted values for the linear model
plot(lm_model$fitted.values, resid(lm_model), 
     xlab = "Fitted Values", ylab = "Residuals", 
     main = "Residuals vs. Fitted Plot")
abline(h = 0, col = "red")  # Horizontal line at 0

```{r}
gamm_model <-gam(tbi ~ s(year) + s(pro_cleanfuel) + s(hdi) + ti(pro_cleanfuel, hdi) + s(country, bs="re"), 
    family = nb(), method = "REML", data = asia_africa)

```{r}
library(car)
vif(gamm_model)
```

```{r}
cor_matrix <- cor(asia_africa[, c("year", "pro_cleanfuel", "hdi", "gdp")], use="pairwise.complete.obs")
print(cor_matrix)

### Longitudinal analysis with Mixed-effect models
```{r}
# Pastikan tidak ada NA dalam variabel yang dianalisis
asia_africa_clean <- asia_africa %>%
  drop_na(tbi, pro_cleanfuel, mean_exposure_pm2.5, pro_basicwater, pro_safewater, pro_basicsanitation, pro_safesanitation, total_greenhouses, gdp, hdi, year, country)

# Model Mixed-Effects: TBI is predicted by predictors with random effect per each country 
model_lmm <- lmer(tbi ~ pro_cleanfuel+ mean_exposure_pm2.5 + pro_basicwater + pro_safewater + pro_basicsanitation + pro_safesanitation + total_greenhouses + gdp + hdi + year + (1 | country),  # Random intercept dressed variation in each country
                   data = asia_africa_clean, REML = FALSE)

# Ringkasan hasil model
summary(model_lmm)
```

@@

# Reshape the dataset to long format
asia_africa_long <- asia_africa %>%
  pivot_longer(cols = c( pro_basicwater, pro_safewater, pro_basicsanitation, 
                        pro_safesanitation,hdi),
               names_to = "Predictor", values_to = "Value")

@@

nonrelationshiplot_location <- here( "code","analysis-code", "figures", "nonlinear.png") # to set up location for the pictures created 
ggsave(filename = nonrelationshiplot_location, plot=nonlinear, width = 12, height = 12, units = "in", dpi = 300) # save the pictures created 
plot(nonlinear)
```

@@

library(dplyr)

asia_africa <- asia_africa %>%
  mutate(
    log_tbi = log(tbi + 1),
    sqrt_basicwater = sqrt(pmax(pro_basicwater, 0)), 
    sqrt_safewater = sqrt(pmax(pro_safewater, 0)),
    poly_basicsanitation = as.numeric(poly(pro_basicsanitation, 2, raw = TRUE)[, 1]),
    poly_safesanitation = as.numeric(poly(pro_safesanitation, 2, raw = TRUE)[, 1])
  )

@@

```{r}
asia_africa_transformed <- asia_africa %>%
  dplyr::select(log_tbi, poly_basicsanitation, poly_safesanitation,
                sqrt_basicwater, sqrt_safewater, hdi) %>%
  pivot_longer(cols = -log_tbi, names_to = "Predictor", values_to = "Value")

colnames(asia_africa)

# Plot after transformation
transformed <- ggplot(asia_africa_transformed, aes(x = Value, y = log_tbi)) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = "lm", color = "red") +  # Use linear model for comparison
  facet_wrap(~ Predictor, scales = "free_x") +
  theme_minimal() +
  labs(title = "Transformed Predictor Relationships with Log(TBI)",
       x = "Transformed Predictor Value", y = "Log(TBI)")

transformedplot_location <- here( "code","analysis-code", "figures", "transformed.png") # to set up location for the pictures created 
ggsave(filename = transformedplot_location, plot=transformed, width = 12, height = 12, units = "in", dpi = 300) # save the pictures created 
plot(transformed)
```

*Generalized Linear mixed model (GLMM)*
After transformation, the data show linear relationship, therefore we perform Linear Mixed Models (LMM) 
```{r}

#Generalized Linear mixed model (GLMM)
#without HDI
library(lme4)
model1<- glmer(log(tbi) ~ sqrt(pro_basicwater) + sqrt(pro_safewater) + 
                          poly(pro_basicsanitation, 2) + poly(pro_safesanitation, 2) +
                          (1 | country), data = asia_africa)
summary(model1)

@@

Interpretation:
✔ Random intercept for country is moderate, meaning tbi varies across countries.
✔ Residual variance is small, indicating that the model explains much of the variability

✔ Intercept (5.24113) is highly significant, representing the baseline log-transformed tbi when all predictors are at their reference values.
✔ poly(pro_basicsanitation, 2)1 (first-order term) is significant, indicating that basic sanitation has a linear effect on tbi.
❌ poly(pro_basicsanitation, 2)2 (second-order term) is not significant, suggesting that the quadratic effect is weak or unnecessary.
❌ pro_basicwater and pro_safewater are not significant, meaning they may not be strong predictors of tbi.

```{r}
#with HDI
library(lme4)
model2<- glmer(log(tbi) ~ sqrt(pro_basicwater) + sqrt(pro_safewater) + 
                          poly(pro_basicsanitation, 2) + poly(pro_safesanitation, 2) +
                          (1 | country) + hdi, data = asia_africa)
summary(model2)

```
Interpretation:

✔ Random intercept for country is moderate, meaning tbi varies across countries.
✔ Residual variance is very small, indicating that the model explains most of the variability, but this may suggest overfitting.

✔ HDI (hdi) is highly significant (p < 0.001), meaning higher HDI is associated with lower tbi.
✔ poly(pro_basicsanitation, 2)1 (first-degree term) is significant, indicating that basic sanitation has a non-linear effect on tbi.
❌ poly(pro_basicsanitation, 2)2 (second-degree term) is not significant, suggesting that the quadratic effect is weak or unnecessary.
❌ pro_basicwater and pro_safewater are not significant, meaning they may not be useful in predicting tbi.

*check wulticolinearity*
```{r}
vif(model_lmm)
```
The Vif test shows that Pro_safesanitation is >5. We need to remove.

*fit model without non-significant predictor*

```{r}
#refit model without non-significant predictor using lmm 
model3 <- glmer(log(tbi) ~ hdi + poly(pro_basicsanitation, 2) + (1 | country), 
                     data = asia_africa)
summary(model3)

```
Interpretation:

The country-level variance (0.42162) is relatively high, indicating that TB incidence varies significantly between countries.
The residual variance (0.01748) is much smaller, meaning a significant portion of TB variation is explained by country-level differences.

Key Findings
✅ HDI is highly significant (Estimate = -3.67, p < 2e-16)

Higher HDI is strongly associated with lower TB incidence.
A 1-unit increase in HDI decreases log(TBI) by ~3.67.
✅ Second-degree polynomial term for sanitation (pro_basicsanitation²) is highly significant (p = 4.15e-06)

Suggests a nonlinear relationship between basic sanitation and TB incidence.
Diminishing or increasing effects at different levels of sanitation.
❌ First-degree polynomial term for sanitation (pro_basicsanitation¹) is NOT significant (p = 0.286)

This indicates that a linear effect of sanitation is weak.
The nonlinear (quadratic) effect is more important.

*Performing Generalized Additive Models (GAM)*

```{r}
model4 <- gamm(log(tbi) ~ s(hdi, bs="ps") + s(pro_basicwater, bs="ps") +
                         s(pro_safewater, bs="ps") + s(pro_basicsanitation, bs="ps") +
                         s(country, bs="re"),
              data = asia_africa_clean, method = "REML")
# bam() from mgcv, which is more efficient for mixed models
summary(model4$gam)
plot(model4$gam)

@@

BIC(model1)
BIC(model2)
BIC(model3)
BIC(model4)
```
```{r}
# Create a data frame for model comparison
model_comparison <- data.frame(
  Model = c("Model 1: Main Predictors", "Model 2: Including Covariates", "Model 3: Significant variables only", "Model 4: GAMM"),
  AIC = c(AIC(model1), AIC(model2), AIC(model3), AIC(model4)),
  BIC = c(BIC(model1), BIC(model2), BIC(model3), BIC(model4))
)

@@

gam_model <- gam(log_tbi ~ s(log_gdp) + s(log_total_greenhouses) + s(log_pm2.5) +
                      s(poly_cleanfuel) + s(poly_basicsanitation) + s(poly_safesanitation) +
                      s(sqrt_basicwater) + s(sqrt_safewater) + s(hdi),
                 data = asia_africa, family = gaussian)
summary(gam_model)

>>>>

```{r}
#load needed packages. make sure they are installed.
library(here)
library(dplyr)
library(skimr)
library(ggplot2)
#install.packages("maps")
library(maps)
#install.packages("ggcorrplot")
library(ggcorrplot)
#install.packages("plm")
library(plm)
#install.packages("gt")
library(gt)
#install.packages("car")
library(car)
#install.packages("MASS")
library(MASS)
install.packages("tidyverse")
library(tidyverse)
#install.packages("reshape2") 
library(reshape2) 
#install.packages("scales")
library(scales)
library(splines)
library(here)
library(knitr)
library(kableExtra)
#install.packages("GGally")
library(GGally)
#install.packages("patchwork")
library(patchwork)
#install.packages("gridExtra")
library(gridExtra)
#install.packages("lmerTest")
library(lmerTest)
#install.packages("stargazer")
library(stargazer)
#install.packages("lmtest")
library(lmtest)
#install.packages("mgcv") 
library(mgcv)  
install.packages("gamm4")
library(gamm4)
install.packages("glmnet")
library(glmnet)
install.packages("brms")
library(brms)
```

@@

seaadf_location <- here("data", "processed-data", "seaa.rds")
seaa_df <- readRDS(seaadf_location)

str(seaa_df)

seaa_df <- seaa_df %>%
  mutate(across(c(pro_basicwater, pro_safewater, 
                  pro_basicsanitation, pro_safesanitation), 
                ~ as.numeric(.)))
```

@@

First, we perform linear regression. Due to high multicolinearity, percentage of population using safe sanitation is eliminated from the model. 
```{r}
# Fit a linear model (without Interaction)
model_lm1 <- lm(tbi ~ pro_basicwater + pro_safewater + 
               pro_basicsanitation +  hdi, 
               data = seaa_df)
summary(model_lm1)
```
```{r}
# Fit a linear model (with Interaction)
model_lm_inte <- lm(tbi ~ (pro_basicwater + pro_safewater + 
               pro_basicsanitation +  hdi)^2, 
               data = seaa_df)
summary(model_lm_inte )
```

```{r}
# Plot residuals vs. fitted values for the linear model
plot(model_lm1$fitted.values, resid(model_lm1), 
     xlab = "Fitted Values", ylab = "Residuals", 
     main = "Residuals vs. Fitted Plot")
abline(h = 0, col = "red")  # Horizontal line at 0
```

# Fit a non-linear model (GAM)
```{r}

model_gam1 <- gam(tbi ~ s(pro_basicwater) + s(pro_safewater) + s(pro_basicsanitation) + s(hdi), 
                 data = seaa_df, method = "REML")

@@

compared the linear model and non linear model using AIC

@@

```{r}
model_gam_inte <- gam(tbi ~ 
                    s(pro_basicwater) + 
                    s(pro_safewater) + 
                    s(pro_basicsanitation) + 
                    s(hdi) + 
                    te(pro_basicwater, pro_safewater) +
                    te(pro_basicwater, pro_basicsanitation) +
                    te(pro_basicwater, hdi) +
                    te(pro_safewater, pro_basicsanitation) +
                    te(pro_safewater, hdi) +
                    te(pro_basicsanitation, hdi),
                  data = seaa_df, method = "REML")
summary(model_gam_inte)
```
```{r}
# Compare AIC values
AIC(model_lm1, model_lm_inte, model_gam1, model_gam_inte)
```
It can be seen that GAM model with two way interaction performs best.

### Longitudinal analysis with Mixed-effect models

```{r}
# make sure to deal with NA data
seaa_df <- seaa_df %>%
  drop_na(tbi,  pro_basicwater, pro_safewater, pro_basicsanitation, pro_safesanitation, hdi, year, country)

# Model Mixed-Effects: TBI is predicted by predictors with random effect per each country 
model_me <- lmer(tbi ~ pro_basicwater + pro_safewater + pro_basicsanitation + pro_safesanitation + hdi + year + (1 | country),  # Random intercept dressed variation in each country
                   data = seaa_df, REML = FALSE)

# Ringkasan hasil model
summary(model_me)
```

@@

# Reshape the dataset to long format
asia_africa_long <- seaa_df %>%
  pivot_longer(cols = c( pro_basicwater, pro_safewater, pro_basicsanitation, 
                        pro_safesanitation,hdi),
               names_to = "Predictor", values_to = "Value")

@@

nonrelationshiplot_location <- here( "results", "figures", "analysis", "nonlinear.png") # to set up location for the pictures created 
ggsave(filename = nonrelationshiplot_location, plot=nonlinear, width = 12, height = 12, units = "in", dpi = 300) # save the pictures created 
plot(nonlinear)
```

@@

seaa_df<- seaa_df %>%
  mutate(
    log_tbi = log(tbi + 1),
    sqrt_basicwater = sqrt(pmax(pro_basicwater, 0)), 
    sqrt_safewater = sqrt(pmax(pro_safewater, 0)),
    poly_basicsanitation = as.numeric(poly(pro_basicsanitation, 2, raw = TRUE)[, 1]),
    poly_safesanitation = as.numeric(poly(pro_safesanitation, 2, raw = TRUE)[, 1])
  )

@@

Visualize the linearity test after transfomation
```{r}
seaa_transformed <- seaa_df %>%
  dplyr::select(log_tbi, poly_basicsanitation, poly_safesanitation,
                sqrt_basicwater, sqrt_safewater, hdi) %>%
  pivot_longer(cols = -log_tbi, names_to = "Predictor", values_to = "Value")

colnames(seaa_transformed)

# Plot after transformation
plot_transformed <- ggplot(seaa_transformed, aes(x = Value, y = log_tbi)) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = "loess", color = "red") +  # Use linear model for comparison
  facet_wrap(~ Predictor, scales = "free_x") +
  theme_minimal() +
  labs(title = "Transformed Predictor Relationships with Log(TBI)",
       x = "Transformed Predictor Value", y = "Log(TBI)")

transformedplot_location <- here("results", "figures",  "analysis", "plot_transformed.png") # to set up location for the pictures created 
ggsave(filename = transformedplot_location, plot=plot_transformed, width = 12, height = 12, units = "in", dpi = 300) # save the pictures created 
plot(plot_transformed)
```
*Performing Generalized Additive Models (GAM)*
```{r}

#  Fit the Bayesian GAMM
model_brms_full <- brm(
  formula = tbi ~ 
    s(pro_basicwater) + 
    s(pro_safewater) + 
    s(pro_basicsanitation) + 
    s(pro_safesanitation) + 
    s(hdi) + 
    s(year) + 
    (1 | country),  # random intercept for longitudinal structure

  data = seaa_df,
  family = negbinomial())

@@

```{r}
# Model summary
summary(model_brms_full)

# Check model convergence and diagnostics
plot(model_brms_full)           # smooth terms & convergence
pp_check(model_brms_full)       # posterior predictive check
conditional_effects(model_brms_full)  # visualize effects

# Check random effects
ranef(model_brms_full)

# Save fitted values or residuals if needed
seaa_df$fitted_tbi <- fitted(model_brms_full)[, "Estimate"]

```{r}
model_brms_full <- brms(
  formula = tbi ~ 
    s(pro_basicwater) + 
    s(pro_safewater) + 
    s(pro_basicsanitation) + 
    s(pro_safesanitation) + 
    s(hdi) + 
    s(year) + 
    (1 | country),

  data = seaa_df,
  family = negbinomial(),

  control = list(adapt_delta = 0.99),  # Fix for divergent transitions

  chains = 4,
  cores = 4,
  iter = 4000
)

)

wash_vars <- seaa_df[, c("pro_basicwater", "pro_safewater", "pro_basicsanitation", "pro_safesanitation")]
pca_result <- prcomp(wash_vars, scale. = TRUE)

# Add first component to your data
seaa_df$PC1 <- pca_result$x[, 1]

model <- gamm4(
  tbi ~ s(PC1) + s(hdi) + s(year),
  random = ~(1 | country),
  family = nb(),  
  data = seaa_df
)

# Define the model formula with smooth terms
formula <- tbi ~ s(hdi) + s(pro_basicwater) + s(pro_safewater) + s(pro_basicsanitation) + s(pro_safesanitation) + s(year)

# Fit the GAMM with a negative binomial family to account for overdispersion
gamm_model <- gamm(formula, random = list(country = ~1), family = nb(), data = seaa_df)

# Extract the design matrix and response variable
X <- model.matrix(gamm_model$gam)
y <- seaa_df$tbi

# Apply elastic net regularization (alpha = 0.5 for elastic net, 0 for ridge, 1 for lasso)
cv_fit <- cv.glmnet(X, y, alpha = 0.5, family = "poisson")

# Obtain the coefficients at the optimal lambda
coef(cv_fit, s = "lambda.min")

after performing data transformation, the data is stil not linear, therefore, Geralized additive mixed morel, which can handle non linearity and account mixed effect of the ciuntries.

```{r}
#convert country into factors

seaa_df$country <- as.factor(seaa_df$country)

model_gamm <- gamm(
  tbi ~ 
    s(hdi) + 
    s(pro_safewater) + 
    s(pro_basicsanitation),
  random = list(country = ~1),
  data = seaa_df,
  method = "REML"
)

model_gamm4 <- gamm4(
  log_tbi ~ 
    s(hdi) +
    s(sqrt_basicwater) +
    s(sqrt_safewater) +
    s(poly_basicsanitation) +
    s(poly_safesanitation),
  random = ~(1 | country),
  data = seaa_df
)

summary(model_gamm4$gam)  # for fixed effects (smooth terms)
summary(model_gamm4$mer)  # for random effects
plot(model_gamm4$gam, pages = 1)

summary(model_gamm_trans$gam)
plot(model_gamm_trans$gam)

@@

BIC(model1)
BIC(model2)
BIC(model3)
BIC(model4)
```

```{r}
# Create a data frame for model comparison
model_comparison <- data.frame(
  Model = c("Model 1: Main Predictors", "Model 2: Including Covariates", "Model 3: Significant variables only", "Model 4: GAMM"),
  AIC = c(AIC(model1), AIC(model2), AIC(model3), AIC(model4)),
  BIC = c(BIC(model1), BIC(model2), BIC(model3), BIC(model4))
)

@@

gam_model <- gam(log_tbi ~ s(poly_basicsanitation) + s(poly_safesanitation) +
                      s(sqrt_basicwater) + s(sqrt_safewater) + s(hdi),
                 data = seaa_df, family = gaussian)
summary(gam_model)
